---
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{amsthm}
title: "2. SAS Data (PCA)"
author: "Hair Parra"
date: "`r Sys.Date()`"
geometry: margin=1.2cm
output: 
    pdf_document: 
      extra_dependencies: ["array", "amsmath","booktabs"]
---

\newtheorem{assumption}{Assumption}[assumption]
\newtheorem{theorem}{Theorem}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}
\newtheorem{example}{Example}
\newtheorem{remark*}{Remark}
\newtheorem{exercise*}{Exercise}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width=9, fig.height=6) 

# configurations for plot 
my_plot_hook <- function(x, options)
  paste("\n", knitr::hook_plot_tex(x, options), "\n")
knitr::knit_hooks$set(plot = my_plot_hook)
```

# SAS Example 

In a survey, 200 adults were asked the following questions regarding a certain type of store.

The responses are provided on a 5 points Likert scale where:
- 1 = not important
- 2 = slightly important
- 3 = moderately important
- 4 = important
- 5 = very important

For you, how important is it:
  - $X_1$ that the store offers good prices every day?
  - $X_2$ that the store accepts major credit cards?
  - $X_3$ that the store offers quality products?
  - $X_4$ that sellers are familiar with the products?
  - $X_5$ that there are special sales regularly?
  - $X_6$ that known brands are available?
  - $X_7$ that the store has its own credit card?
  - $X_8$ that the service is fast?
  - $X_9$ that there is a wide selection of products?
  - $X_{10}$ that the store accepts payment by debit card?
  - $X_{11}$ that the staff is courteous?
  - $X_{12}$ that the store has in stock the advertised products?



```{r}
# load required libraries 
library("here") 
library("haven")

# Load the 'haven' library for importing SAS data files
library(haven)

# Read the data from a SAS file and store it in the 'datpca' variable
datpca <- read_sas(here("code_data_W2024", "factor2.sas7bdat"))

# Convert the data to a matrix
datpca <- as.matrix(datpca)

# Standardize (scale) each column of the matrix
datpca <- apply(datpca, 2, scale) # i.e. mean 0, std 1 
```

## PCA 

Apply ordinary PCA and get a **scree_plot**. 

```{r}
# Load the 'factoextra' library for visualizing PCA results
library(factoextra)

# Perform Principal Component Analysis (PCA) on the standardized data 'datpca'
pca <- prcomp(datpca)

# Get and print the eigenvalues, variance percentages, and cumulative variance percentages
get_eig(pca)
``` 

```{r}
# Visualize the eigenvalues using a scree plot
fviz_eig(pca)
```

- Three or four components seem reasonable according to the scree plot. 
- The first 4 recover 64% of the variance in the data. 

```{r}
# getting the coefficient of the first four components
pca[[2]][,1:4]
```
- All the variables contribute (i.e. have a non-zero coeficients) to all components.
- **However we need interpretation!** 
e.g. **factor analysis** 
- Can us ea **rotation method**: solution with many small *loadings* and only a few large ones. 
- Easier to interpret is mabhy coefficients were 0 $\implies$ **Lasso regularization can be helpful!**. 

## Lasso PCA 

```{r}
# load required libraries 
library(elasticnet)

# Perform Sparse PCA on the standardized data 'datpca'
spcafit=spca(datpca,
             K=4, # number of components
             sparse="penalty", # penalty method
             para=rep(0,4)) # values of lambda for each component
```

**Note:** We set all the lambdas to zero for no shrinkage. 

```{r}
spcafit
```

If we use a $\lambda$ of 32 for each component, we get the following sparse solution:

```{r}
# Perform Sparse PCA on the standardized data 'datpca'
spcafit=spca(datpca, K=4, sparse="penalty", para=rep(32,4))
spcafit
```
This is the solution that we get from a **classical factor analysis**. 

- The first component is related to **service** 
- The second one to **products** 
- The third one to **payment** 
- The fourth one to **price** 

**Note:** Can use a genetic algorithm for hyperparameter selection. 



